---
title: "Project Proposal"
author: "Nilutpaul Sarker Yash, Grace Schmitt, Sarah Bass, Kate Passwater"
output: html_document
---

# Research Topic

Healthcare Insurance and Factors Affecting the Cost of Insurance for a Person


# Team Members

- Nilutpaul Sarker Yash
- Grace Schmitt
- Sarah Bass
- Kate Passwater

# Dataset

We are using the following dataset: https://www.kaggle.com/datasets/willianoliveiragibin/healthcare-insurance?select=insurance.csv

```{r setup, include=FALSE}
library(tidyverse)

insurance <- read_csv("https://raw.githubusercontent.com/stedy/Machine-Learning-with-R-datasets/master/insurance.csv")
```

# Data Description

The dataset contains information on the insurance costs of individuals and their personal attributes, such as age, sex, BMI, number of children, smoking status, and region. The initial data cleaning steps we expect at this point are checking the variable types, looking for missing values, and separating each factor with insurance cost to analyze each trend.

# Project Idea / Research Questions

We have thought of 6 potential research questions from the data to this point. They include:

a)	Age and Charges - Is there a relationship between age and insurance charges?

b)	Sex and Charges – Do insurance charges differ by sex?

c)	BMI and Charges - Is there a relationship between BMI and insurance charges?

d)	Smoker and Charges - Does smoking increase insurance charges?

```{r}
ggplot(insurance, aes(x = smoker, y = charges, fill = smoker)) +
  geom_boxplot() +
  labs(title = "Charges by Smoking Status")
```
#The boxplot above shows a very strong correlation with charges related to smokers vs non smokers. We can tell this because the median line for the non-smokers (red) is much lower than that of smokers (blue). A typical smoker costs around $30,000-$40,000 while non-smokers are around $5,000-$10,000. There is obviously a tiny bit of overlap between smokers and non-smokers because of other health factors. However, those people are considered outliers and are not common. 



g) Age and Smoking- are older smokers charged more than those younger
 
h) Smoking and region- Do smokers in certain regions get charged more?

```{r}
ggplot(insurance, aes(x = region, y = charges, fill = smoker)) +
  geom_boxplot() +
  labs(title = "Insurance Charges by Region and Smoking Status",
       x = "Region",
       y = "Charges") +
  theme_minimal()
```
#The boxplot shows that smoking status has a significantly larger impact on insurance charges than geographic region. In every region, smokers consistently have much higher median charges compared to non-smokers. While there are slight differences across regions (with the southeast and southwest appearing marginally higher), these differences are small relative to the gap between smokers and non-smokers. This suggests that smoking is a primary driver of insurance cost, while regional variation plays a minor role. Additionally, smokers exhibit greater variability in charges, indicating that other risk factors may further influence pricing within this group.

e)	Children and Charges - Does insurance cost change with the number of dependents?

f)	Region and Charges - Are insurance charges greater in certain regions?


# Planned Analyses / Visualizations

Initially, we plan to use simple visualizations to observe trends in the data. We expect to use scatterplots, boxplots, and bar charts to compare variables and better understand patterns or possible outliers.