---
title: "ProjectCombined_rmd"
author: Nilutpaul Sarker Yash, Sarah Bass, Grace Schmitt, Kate Passwater
output: html_document
date: "2026-05-05"
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)

insurance <- read.csv("insurance.csv")
dat <- insurance
```

# Yash's Section

```{r}
library(tidyverse)

insurance <- read.csv("insurance.csv")

# Check structure
glimpse(insurance)

# Convert categorical variables to factors
insurance <- insurance %>%
  mutate(
    sex = factor(sex),
    smoker = factor(smoker),
    region = factor(region)
  )

insurance <- insurance %>%
  mutate(
    sex = factor(sex),
    smoker = factor(smoker),
    region = factor(region)
  )

glimpse(insurance)
```

```{r}
insurance %>%
  ggplot(aes(x = sex, y = charges, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Insurance Charges by Sex",
    x = "Sex",
    y = "Insurance Charges"
  ) +
  theme_light(base_size = 16) +
  theme(legend.position = "none")
```

- The boxplot comparing insurance charges by sex shows that males tend to have a slightly higher median insurance charge than females. Additionally, the variability in charges appears greater for males, as indicated by a wider spread of the data. Both groups feature multiple high-cost outliers, suggesting that some individuals face significantly higher expenses regardless of sex. Overall, sex is not a strong predictor of insurance charges because the distributions for males and females largely overlap, and the variability within each group is high, meaning that knowing a person’s sex does not provide much information about their expected insurance cost.

```{r}
insurance %>%
  ggplot(aes(x = bmi, y = charges, color = bmi)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient(low = "#56B1F7", high = "#132B43") +
  labs(
    title = "BMI vs Insurance Charges",
    x = "BMI",
    y = "Charges",
    color = "BMI"
  ) +
  theme_light(base_size = 16)
```

```{r}
insurance <- insurance %>%
  mutate(bmi_group = ifelse(bmi >= 30, "BMI ≥ 30", "BMI < 30"))

insurance %>%
  ggplot(aes(x = bmi_group, y = charges, fill = bmi_group)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Insurance Charges by BMI Category",
    x = "BMI Category",
    y = "Charges"
  ) +
  theme_light(base_size = 16) +
  theme(legend.position = "none")
```

- The scatterplot of BMI versus insurance charges was used to explore the relationship between the two variables and shows a weak to moderate positive trend, where higher BMI values are generally associated with higher insurance charges. However, there is substantial variability across all BMI levels, indicating that BMI alone is not a strong predictor of insurance cost. While examining the plot, I noticed a clustering of higher charges among individuals with BMI values around and above 30, suggesting a potential threshold effect.

- To further investigate this observation, I grouped the data into two categories (individuals with BMI < 30 and individuals with BMI ≥ 30) and visualized the distributions using a boxplot. The boxplot confirms that individuals with BMI ≥ 30 tend to have higher median charges, greater variability, and more extreme high-cost values. Overall, while BMI does not strongly predict insurance charges in a strictly linear sense, it exhibits a clear threshold effect beyond a BMI of 30, where costs increase more sharply.

Sarah's Section:

```{r}
#read and inspect the data
dat <- read.csv("insurance.csv")

head(dat)
glimpse(dat)

#check for NA values in each category in case we need to fix anything
sum(is.na(dat$age))
sum(is.na(dat$sex))
sum(is.na(dat$bmi))
sum(is.na(dat$children))
sum(is.na(dat$smoker))
sum(is.na(dat$region))
sum(is.na(dat$charges))

library(ggplot2)
library(corrplot)


#change data into factors for a correlation matrix
dat<- dat %>%
  mutate(sex= factor(sex),
         smoker= factor(smoker),
         region= factor(region))

#change from factor with characters to factor with numerical values
dat<- dat %>%
  mutate(region= case_when(
        region== "northwest" ~1,
        region== "northeast" ~2,
        region== "southwest" ~3,
        region== "southeast" ~4
         ),
        sex= case_when(
          sex== "female" ~0,
          sex== "male" ~1),
        smoker= case_when(
          smoker== "yes" ~1,
          smoker== "no" ~0
        ))

#create the correlation matrix and plot
cor_matrix <- cor(dat)
corrplot(cor_matrix, method = "color", type = "full",
         col= colorRampPalette(c("#ADD8E6", "white", "#FFA07A"))(200),
         tl.col = "black", tl.srt = 45, # text label color & rotation
         addCoef.col = "black",         # add correlation coefficients
         number.cex = 0.7)              
```

```{r}
#for e
library(ggthemes)
ggplot(dat, aes(x=factor(children), y=charges, fill=children)) +
  geom_boxplot() +
  labs( title= "Charges vs Number of Children",
        x= "Number of Children",
        y= "Charges ($)") + 
  theme_light()
  
```

# The median charges seem to increase as the number of children increase, up until 5, where the median charges drop. From 3-5 children, the range decreases. For 0-1 children, the range and IQR of charges are similar. So charges typically increase as number of children increase, until 5 children, where charges were much less than charges for families with less children.

```{r}
#for g
library(ggthemes)
smoke_dat<- dat %>%
  filter(smoker== "yes")

ggplot(dat, aes(x=age, y= charges, color= factor(age))) +
  geom_point() +
  facet_wrap(~smoker)+
  theme_few() +
  labs( title= "Charges vs. Age by Smoking Status", 
          x= "Age",
        y= "Charges ($)")
```

# it can be seen that non smokers have lower baseline charges than smokers, and both group's charges increase as age increases.

# Grace's Section:

```{r setup, include=FALSE}
insurance <- read_csv("https://raw.githubusercontent.com/stedy/Machine-Learning-with-R-datasets/master/insurance.csv")

```

```{r}
#graph the distribution of smokers vs non-smokers in this dataset to see if there wil be bias
smoking_stat<- dat %>%
  group_by(smoker) %>%
  summarise(smoke_totals= n()) %>%
  ungroup()

ggplot(smoking_stat,aes(x= smoker, fill=smoker)) +
  geom_bar(aes(weight= smoke_totals)) +
  labs(title= "Distribution of Smokers",
       x= "Smoking Status",
       y="count")
```

```{r}
#graph the relationship

ggplot(insurance, aes(x = smoker, y = charges, fill = smoker)) +
  geom_boxplot() +
  labs(title = "Charges by Smoking Status")
```

# The boxplot above shows a very strong correlation with charges related to smokers vs non smokers. We can tell this because the median line for the non-smokers (red) is much lower than that of smokers (blue). A typical smoker costs around $30,000-$40,000 while non-smokers are around $5,000-$10,000. There is obviously a tiny bit of overlap between smokers and non-smokers because of other health factors. However, those people are considered outliers and are not common. 

```{r}
ggplot(insurance, aes(x = region, y = charges, fill = smoker)) +
  geom_boxplot() +
  labs(title = "Insurance Charges by Region and Smoking Status",
       x = "Region",
       y = "Charges") +
  theme_minimal()
```

# The boxplot shows that smoking status has a significantly larger impact on insurance charges than geographic region. In every region, smokers consistently have much higher median charges compared to non-smokers. While there are slight differences across regions (with the southeast and southwest appearing marginally higher), these differences are small relative to the gap between smokers and non-smokers. This suggests that smoking is a primary driver of insurance cost, while regional variation plays a minor role. Additionally, smokers exhibit greater variability in charges, indicating that other risk factors may further influence pricing within this group.

# Kate's Section:

```{r}
library(ggplot2)
data <- read.csv("insurance.csv")
data
```

#Age and Charges
```{r}
#to get a regression line
correlation<- cor(data$age, data$charges)


data |> ggplot(aes(x=age, y=charges, color=age))+geom_point(alpha=0.5)+
  labs(title="Relationship between Age and Insurance Charges", x="Age", y="Insurance Charges") +
  geom_smooth(method=lm,, se= FALSE, col = "red", lwd = 1) +
  annotate("text", x=-Inf, y=Inf, label= paste("r =", round(correlation, 2)),
            hjust=-0.5, vjust=1.5, color="red", fontface= "bold")
  
```


```{r}
#get the correlations for each smoking group
cor_no<- cor(dat$age[dat$smoker== "no"], dat$charges[dat$smoker=="no"])
cor_yes<- cor(dat$age[dat$smoker== "yes"], dat$charges[dat$smoker=="yes"])

#graph to show difference in charges with the correlations
data |> ggplot(aes(x=age, y=charges, color=smoker))+geom_point(alpha=0.5)+labs(title="Age and Insurance Charges for Smokers v Non-Smokers", x="Age", y="Insurance Charges") + geom_smooth(method= lm, se= FALSE) +
  scale_color_discrete(labels= c(paste("no, (r = ", round(cor_no,2), ")"),
                                 paste("yes, (r = ", round(cor_yes, 2), ")")))
```

#Region and Charges

```{r}
#graph to show the relationship between region and charges
data |> ggplot(aes(x=factor(region), y=charges, fill=region))+geom_boxplot()+labs(title="Relationship between Region and Insurance Charges", x="Regions", y="Insurance Charges")

```


```{r}
#to show the distribution of regions among the dataset
dat_region<- data %>%
  group_by(region) %>%
  summarise(region_counts= n()) %>%
  ungroup()


ggplot(dat_region,aes(x= region, fill=region)) +
  geom_bar(aes(weight= region_counts)) +
  labs(title= "Distribution of Regions",
       x= "Region",
       y="count")
```